Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-12-27 paper/paper.tex
partially added text about stochasticity to last results section

@@

To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden among cells.

Cells infected by incomplete or mutated virions also expressed IFN more frequently than cells infected by full wild-type virions (\FIG{mutations}B), although this difference was not statistically significant ($P = 0.12$, Fisher's exact test).
However, some specific viral defects were significantly associated with IFN induction: absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were weakly enriched (\FIG{mutations}C).
Due to the low number of cells and large number of hypotheses in \FIG{mutations}C, the only trend that remained significant at a false discovery rate of 10\% was absence of NS.

One other interesting trend emerges from the single-cell data.
There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).

\subsection*{Validation that viral defects in single IFN+ cells often cause increased IFN induction}

@@

\centerline{\includegraphics[width=0.85\textwidth]{figures/Validation_Figure/ifn_plot.pdf}}
\caption{
{\bf Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.}
{\bf (A)}
Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
The numbers indicate the median of four measurements for each viral mutant.
The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wild type is indicated with a dashed blue line.
Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
{\bf (B)}
Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
}
\label{fig:validation}

@@

To test if the viral defects identified in single IFN+ cells play a causal role in triggering IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.

The viral defect most strongly associated with IFN induction was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.

The single-cell data also showed that amino-acid substitutions in the PB1 and NS genes were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with many of the amino-acid substitutions found in these genes among IFN+ cells.
Overall, we created six such point-mutant viruses: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.

Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
We therefore created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.

We tested the rate of IFN induction by each viral stock using the reporter cells.
\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wild type.
The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we also expect some mutations without an IFN-enhancing effect to be found in IFN+ cells by chance.
But overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.

>>>>

\newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}

@@

To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden among cells.

Cells infected by incomplete or mutated virions also expressed IFN more frequently than cells infected by full wild-type virions (\FIG{mutations}B), although this difference was not statistically significant ($P = 0.12$, Fisher's exact test).
However, some specific viral defects were significantly associated with IFN induction: absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were weakly enriched (\FIG{mutations}C).
Due to the low number of cells and large number of hypotheses in \FIG{mutations}C, the only trend that remained significant at a false discovery rate of 10\% was absence of NS.
However, the validation experiments in the next section show that the lack of statistical significance for many of the trends is due to the modest number of infected cells that were sequenced rather than a lack of true association between viral defects and IFN induction.

One other interesting trend emerges from the single-cell data.
There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS (\FIG{mutations}D).
But among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
\jdbcomment{Perhaps elaborate a bit after adding additional data relevant to this point to next section.}

\subsection*{Validation that viral defects in single IFN+ cells often increase IFN induction}

@@

\centerline{\includegraphics[width=0.85\textwidth]{figures/Validation_Figure/ifn_plot.pdf}}
\caption{
{\bf Validation that IFN induction is increased by mutations identified in the single-cell virus sequencing of IFN+ cells.}
{\bf (A)}
Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated mutant, as determined using a reporter cell line.
Numbers give the median of four measurements for each mutant.
The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wild type is indicated with a dashed blue line.
Points are orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
{\bf (B)}
Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
The experiments in the two panels were performed on different days, and so numerical values can be compared within but not between panels.
}
\label{fig:validation}

@@

To test if the viral defects identified in single IFN+ cells directly increase IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.

The viral defect most strongly associated with IFN induction was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
We therefore mimicked the effect of absence of NS by creating a mutant virus (NS1stop) that had multiple stop codons early in the NS1 coding sequence.

The single-cell data also showed that amino-acid substitutions in PB1 and NS were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with some of the substitutions found in these genes among IFN+ cells: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.

Finally, prior work has suggested that virions with internal deletions in the polymerase genes can induce IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is a co-infected IFN+ cell where one viral variants has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
We created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.

We tested the rate of IFN induction by each viral stock using the reporter cells.
\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than a wild type viral stock.
The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we also expect some mutations without an IFN-enhancing effect to be found in IFN+ cells by chance.
But overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.

However, IFN induction remains stochastic even for the most strongly IFN-inducing viral mutants.
\FIG{validation} shows flow cytometry data (see also \SUPPFIG{validationflow}), which is itself a single-cell measurement, albeit one that does not report the viral genotype.
All the viral mutants induce IFN in less than a fifth of cells
